---
title: "Data622_Assignment_2"
author: "Natalie Kalukeerthie and Anna Moy"
date: "2025-09-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1

```{r,message= FALSE, warning=FALSE}
library(tidyverse) 
library(ggplot2)
library(corrplot)
library(dplyr)

#import full bank dataset

# Import the CSV file from github
bank <- read_csv2("https://raw.githubusercontent.com/nk014914/Data622/refs/heads/main/bank-full.csv", show_col_types = FALSE)

```

First, we'll look at the structure of the data.

```{r pressure, echo=FALSE}
# Preview structure
str(bank)
summary(bank)

# check for duplicates
sum(duplicated(bank))

```

Our chosen dataset,'Bank-Full', has 45,211 rows and 17 columns. It contains both numerical variables (e.g., age, balance, duration) and categorical variables (e.g., job, marital status, education). Our target variable is y, which indicates whether a client subscribed to a term deposit. Having a large dataset allows us to run models more effectively.

The summary provides central tendency (mean, median) and spread (quartiles, min, max) for numeric variables, and frequency counts for categorical ones. This helps us quickly spot imbalances or unusual values.

1.  Are the features (columns) of your data correlated?

```{r}
numeric_vars <- bank %>% select_if(is.numeric)
corr_matrix <- cor(numeric_vars)
corrplot(corr_matrix, method = "color", addCoef.col = "black")
```

The correlation matrix shows that most numeric features are weakly correlated with each other. For example, duration and campaign have a very low correlation. This indicates that predictors are largely independent, which is good for most modeling approaches. With some indicating zero it means there is no correlation with other features in the data. 

2.  What is the overall distribution of each variable?

```{r}
#numerical
numeric_vars %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  facet_wrap(~Variable, scales = "free") +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  theme_minimal()

#categorical
categorical_data <- select(bank, where(is.character))

categorical_data %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = value)) +
  facet_wrap(~name, scales = "free", ncol = 3) +
  geom_bar(fill = "darkorange")
```

Numerical data: We see that age is right-skewed, with most clients being 25–60. Balance is highly skewed with extreme outliers, along with campaign and duration also being skewed.

Categorical data: There is imbalance of data as the subscribe a term deposit has higher numbers in no than yes. The distribution of the features are varying in distributions whcih some higher than others. 

3.  Are there any outliers present?

```{r}
boxplot(bank$balance, main = "Balance Outliers")
boxplot(bank$duration, main = "Duration Outliers")

numeric_vars %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(y = value)) +
  facet_wrap(~name, scales = "free", ncol = 3) +
  geom_boxplot()
```

There does seem to be some outliers present. Balance and duration show extreme outliers that may affect model training.

4.  What are the relationships between different variables?

```{r}
ggplot(bank, aes(x = age, y = balance, color = y)) +
  geom_point(alpha = 0.5) +
  theme_minimal()
```

The scatterplot of age vs. balance shows that subscription decisions are not strictly determined by financial status. Younger clients (20s–30s) often have lower balances and mixed responses, while middle-aged clients (30s–50s) with moderate balances appear somewhat more likely to subscribe, possibly due to financial stability and savings goals. Older clients (60+) frequently have higher balances but show fewer “yes” outcomes, suggesting other priorities. Overall, the relationship is noisy (no single trend is distinct) indicating that age and balance alone cannot reliably predict subscription outcomes.

5.  How are categorical variables distributed?

```{r}
ggplot(bank, aes(x = marital)) + geom_bar(fill = "skyblue")
ggplot(bank, aes(x = education)) + geom_bar(fill = "skyblue")
ggplot(bank, aes(x = job)) + geom_bar(fill = "skyblue")
ggplot(bank, aes(x = y)) + geom_bar(fill = "darkgreen")

bank %>%
  select(job, marital, education, housing, loan, y) %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = value)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~name, scales = "free_x") +
  coord_flip()

ggplot(bank, aes(x = job, fill = y)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  coord_flip()

ggplot(bank, aes(x = interaction(job, education), fill = y)) +
  geom_bar(position = "fill") +
  coord_flip() +
  ylab("Proportion") +
  xlab("Job & Education Combination") +
  ggtitle("Subscription Rate by Job and Education")


ggplot(bank, aes(x = interaction(marital, loan), fill = y)) +
  geom_bar(position = "fill") +
  coord_flip() +
  ylab("Proportion") +
  xlab("Marital Status & Loan") +
  ggtitle("Subscription Rate by Marital Status and Loan")

```

Most clients are married and have secondary education. Blue-collar, management, and technician are the most common jobs. The target variable y shows a heavy imbalance, with many more “no” responses than “yes.”. There sems to be a correlation of 

6.  Do any patterns or trends emerge in the data?

```{r}
ggplot(bank, aes(x = duration, fill = y)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
  theme_minimal()

ggplot(bank, aes(x = month, fill = y)) +
  geom_bar(position = "fill") +
  ylab("Proportion Subscribed")


```

Yes, we're able to see a few patterns from the data, namely:

-   Call duration is strongly predictive (longer calls are much more likely to end in a “yes.”)

-   Middle-aged clients with stable jobs and balances appear somewhat more responsive.

-   Campaign outcome is imbalanced, meaning the bank struggled to achieve a high conversion rate overall.

-  There were more yes in the month of december, march, october and september than other months. 

7.  What is the central tendency and spread of each variable?

```{r}
summary(numeric_vars)
```

The central tendency and spread of the numeric variables provide insights into the typical client profile as well as the variability in the data.

-   `age`: mean about 40 years, range 18–95.

The age variable has a mean of approximately 40 years, with clients ranging from 18 to 95 years old, indicating that the dataset includes a wide age spectrum, though most clients fall within the 25–60 range.

-   `balance`: median near 448 but very high max (\>80,000) showing extreme outliers.

The balance variable is heavily skewed: while the median balance is around 448, some clients have extremely high balances exceeding 80,000, revealing the presence of significant outliers that could influence model performance.

-   `duration`: median \~180s, but some calls last over an hour.

Similarly, duration, which measures the length of the last call in seconds, has a median value around 180 seconds, but a few calls exceed an hour, again highlighting extreme cases.

-   `campaign`: median 2 contacts, max 63 (also extreme outliers).

The campaign variable, representing the number of contacts performed during this campaign for a client, has a median of 2 and a maximum of 63, showing that some clients were contacted repeatedly. Overall, these statistics indicate that most clients cluster around typical values (middle-aged, moderate balances, short calls, and few campaign contacts), but the presence of extreme outliers and wide spreads suggests that the dataset has heavy tails and variability that should be considered during modeling.

8.  Are there any missing values and how significant are they?

```{r}
summary(numeric_vars)
print(bank)
```

The dataset contains no missing values, so no imputation is required.

# Assignment #2

To prepare out model for training we want to remove the duration of the calls the customer had. By removing this it will avoid the model from looking at them to get the result. We also added a column for easy visibility to find out if we were able to contact the client or not. Later on we split the training and test data into a 80 and 20 split. 

```{r, message= FALSE, warning=FALSE}
library(tidymodels)
library(caret)

# Drop duration (to avoid data leakage on how long the call lasted with the customer)
bank <- bank %>% select(-duration)

# Create new column called pdays_flag to indicate if the pdays are contacted or never contacted
bank <- bank %>%
  mutate(
    pdays_flag = if_else(pdays == -1, "Never Contacted", "Contacted"),
    pdays = if_else(pdays == -1, NA_real_, pdays)  # Optional
  )

# Convert character variables to factors
bank <- bank %>%
  mutate(across(where(is.character), as.factor))

# Train-test split (80/20)
set.seed(123)
split <- initial_split(bank, prop = 0.8, strata = y)
train_data <- training(split)
test_data <- testing(split)

```


We are adding performance metrics for us to check later on how each model is performing. The metrics being used are:
- Accuracy - Predicts how well our model is performing overall
- Precision - For those labeled as positive how many are actually positive
- Recall - Of all the actual positives. How many did we find
- F1-Score - This is the balance between precision and recalls. We want to have a balance between false positives and false negatives. 


```{r}
library(yardstick)

my_metrics <- metric_set( accuracy, yardstick::precision, yardstick::recall, f_meas )

```

The first model we implemented is the baseline Decision Tree. This serves as a reference point for comparing any modifications we make to the tree structure and parameters in our next model.

The results from the baseline model show strong performance. It achieved an accuracy of 89.3%, indicating that it correctly classified a high proportion of the test data overall. Additionally, the precision, recall, and F1-score are all above 90%, suggesting the model is effective at identifying positive cases, with few false positives and a good balance between precision and recall.

```{r}
#Decision Tree (Baseline)
library(rpart)
library(rpart.plot)

# Model specification
dt_model <- decision_tree(mode = "classification") %>%
  set_engine("rpart")

# Fit model
dt_fit <- dt_model %>%
  fit(y ~ ., data = train_data)

# Predict
dt_preds <- predict(dt_fit, new_data = test_data, type = "prob") %>%
  bind_cols(predict(dt_fit, test_data)) %>%
  bind_cols(test_data %>% select(y))

# Evaluate
# Calculate class-based metrics (accuracy, precision, recall, f1-score)
dt_class_metrics <- my_metrics(
  data = dt_preds,
  truth = y,
  estimate = .pred_class
)

# View results
print(dt_class_metrics)

```

The second model we ran was an experiment to see whether changing the depth of the decision tree would reduce overfitting and potentially improve performance. In this version, we set the tree depth to 3 and the minimum split to 2. We evaluated the model using accuracy, precision, recall, and F1-score, comparing the results to those of the baseline decision tree. The metrics showed that the model performed quite well. To check for overfitting, I compared the evaluation metrics on both the training and test data. Since the results were fairly consistent across both sets, it suggests that the model is not overfitting.

In comparison to the default Decision tree the results of the model show the accuracy and recall is higher in the second model but the precision and f1-score was slightly lower than the baseline model. With a small difference between the two models it can be concluded the model are fairly similar in results even with the changes in depth and split. 


```{r}
#Decision Tree (3 depth and 2 split)
dt_model_deep <- decision_tree(mode = "classification") %>%
  set_engine("rpart", control = rpart.control(maxdepth = 3, minsplit = 2, cp = 0))

dt_fit_deep <- dt_model_deep %>%
  fit(y ~ ., data = train_data)

dt_preds_deep <- predict(dt_fit_deep, new_data = test_data, type = "prob") %>%
  bind_cols(predict(dt_fit_deep, new_data = test_data)) %>%
  bind_cols(test_data %>% select(y))

dt_metrics_deep <- my_metrics(
  data = dt_preds_deep,
  truth = y,
  estimate = .pred_class
)

print(dt_metrics_deep)
train_preds <- predict(dt_fit_deep, new_data = train_data, type = "prob") %>%
  bind_cols(predict(dt_fit_deep, new_data = train_data)) %>%
  bind_cols(train_data %>% select(y))

train_metrics <- my_metrics(
  data = train_preds,
  truth = y,
  estimate = .pred_class
)

print(train_metrics)

```

To better understand the change in parameters in the decision tree we visually showed the difference on how the tree would look like. 

```{r}
# Extract rpart fit object from parsnip fit
rpart_base <- dt_fit$fit
rpart_deep <- dt_fit_deep$fit

library(rpart.plot)

rpart.plot(rpart_base, main = "Decision Tree (Baseline)")
rpart.plot(rpart_deep, main = "Decision Tree (3 depth and 2 minisplit)")

```


The third model we ran was the Random Forest baseline similar to what we did for the decision tree model we want to compare the Random Forest baseline to see how it affects our metric results if we change the parameters. The results of the Random Forest baseline shows the accuracy as 71.9% which is decent but there can be better improvements in the model. The precision is performing great as the result is 91.8%. The recall is 74.9% which shows it is missing some positives and f1-score is 82.5% which shows it has decent balance. 

```{r}
#Random Forest(Baseline)
rf_model <- rand_forest(mode = "classification") %>%
  set_engine("ranger")

rf_fit <- rf_model %>%
  fit(y ~ ., data = train_data)

rf_preds <- predict(rf_fit, new_data = test_data, type = "prob") %>%
  bind_cols(predict(rf_fit, test_data)) %>%
  bind_cols(test_data %>% select(y))

rf_metrics <- my_metrics(
  data = rf_preds,
  truth = y,
  estimate = .pred_class
)

print(rf_metrics)
```

For our parameter changes in the Random Forest we want to increase the number of trees to 500 and set mtry to 4 to see if this will improve the performance of the model compared to the Random Forest baseline model. By increasing the number of trees this should provide better predictions and lead to better generalization and when we change the number of features for each split it can improve performance. The results of this model shows it performed slightly better than the Random Forest model in accuracy, precision, recall and f1-score

```{r}
#Random Forest (increase 500 trees and 4 features)
rf_model_tuned <- rand_forest(mode = "classification", trees = 500, mtry = 4) %>%
  set_engine("ranger")

rf_fit_tuned <- rf_model_tuned %>%
  fit(y ~ ., data = train_data)

rf_preds_tuned <- predict(rf_fit_tuned, new_data = test_data, type = "prob") %>%
  bind_cols(predict(rf_fit_tuned, test_data)) %>%
  bind_cols(test_data %>% select(y))

rf_metrics_tuned <- my_metrics(
  data = rf_preds_tuned,
  truth = y,
  estimate = .pred_class
)
rf_metrics_tuned

```


